Performance comparison of Dask and Apache Spark on HPC systems

### 2022-02-23 latex/paper.tex
[UPDATE] fix few comment and generate gantt chart for new kmeans experiments.

We configured Lustre to have 1 MGS and 4 OSS.
Each OSS has 11 OSTs (disks) of \SI{8.8}{\tebi\byte}.
The OSTs are configured to have a maximum of \SI{1000}{\mebi\byte} of dirty data written into the client page cache.
Both the MDT and OSTs are configured to a maximum of 64 concurrent RPCs in flight.
A timeout of \SI{100}{\second} is set for Lustre, and OSTs checksum are disabled.
Lustre \textit{2.14.0-1.el8} is installed on both the Lustre clients and servers.
%  TODO Inode config??

@@

\subsubsection{Increment}
We adapted the increment application used in~\cite{8752675}.
This synthetic application reads blocks of the BigBrain from Lustre and
simulates computation by sleeping for a specified period (Fig.~\ref{fig:graph-increment}). To simulate
intermediate results, we repeat the sleep process for a configurable amount
of time. We prevent data caching of the blocks by incrementing their voxels
value by one after each sleep operation. Finally, we write the resulting
NIfTI image back to Lustre. This application allows us to study the engines
when their inputs are processed independently. The map-only pattern
mimics the processing of multiple independent subjects in parallel.
% \begin{figure}[!hb]
% 	\centering
% 	\includegraphics[height=0.5\columnwidth,
% 	angle=0]{figures/increment.png}
% 	\caption{Task graph for Incrementation with 3 iterations and 3 BigBrain blocks.}
% 	\label{fig:graph-increment}
% \end{figure}

@@

@@

\begin{table*}[t]
	\renewcommand{\arraystretch}{1.5}
	\caption{Parameters for the experiments}\label{table:parameters}
	\centering
	\begin{tabular}{|l|c|c|c|c|c|}
		\hline           & Increment & Histogram & Kmeans  & BIDS App Example \\\hline
		Dataset size [\SI{}{\gibi\byte}] &\multicolumn{2}{c|}{606} & 303 & \multicolumn{1}{c|}{26.67} \\\hline
		\# of Nodes & \multicolumn{4}{c|}{2, 4, 8} \\\hline
		Number of file & \multicolumn{2}{c|}{1000, 2500, 5000} & - & \multicolumn{1}{c|}{3491} \\\hline
		\# of Iterations & 1, 5, 25  & -         & 1, 3, 9 & -                \\\hline
	\end{tabular}
\end{table*}

@@

From Figure \ref{fig:increment_worker}, we also observe that Dask's overhead is higher than Spark's.
While the overhead of both engines increases with the number of nodes, Dask's overhead is significantly worse when scaling.
Although Dask has a higher overhead, the total time for both engines is similar; with a maximum makespan difference of 2\% for two nodes.
This is due to a tradeoff between disk bandwidth and compute/overhead time.
On the one hand, multiple threads access Lustre concurrently, thus sharing the disk bandwidth resulting in longer I/O time.
On the other hand, when most threads are either idle or performing computations, it reduces the contention on the disk bandwidth resulting in shorter I/O time.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_increment_worker.jpg}
	\caption{Increment: varying nodes -- 5000 files, 5 iterations, \SI{1}{\second} delay}
	\label{fig:increment_worker}
\end{figure}

@@

Figure \ref{fig:increment_itr} also shows that the the total I/O time decreases as total computation time increases.
This is explained by an I/O bottleneck on the Lustre file system.
When threads are busy with computations, it reduces the contention on Lustre, thus improving I/O speed for individual accesses.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_increment_itr.jpg}
	\caption{Increment: varying iterations -- 4 nodes, 5000 files, \SI{1}{\second} delay}
	\label{fig:increment_itr}
\end{figure}

@@

From Figure \ref{fig:increment_block} we also observe that Dask has significantly higher overhead than Spark.
However, the total execution time of the application is similar due to the tradeoff between I/O and compute/overhead time.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_increment_block.jpg}
	\caption{Increment: varying \# of files -- 4 nodes, 5 iterations, \SI{1}{\second} delay}
	\label{fig:increment_block}
\end{figure}

\subsection{Histogram}
Figure \ref{fig:histogram_worker} depicts the execution time for the \textit{Histogram} application when varying the number of nodes.
Overall, Spark's total execution time is up to 5\% lower than Dask's, with the difference coming from a shorter execution time for \textit{Calculate\_histogram}.
This is unexpected as the code base for both engines is the same, with only the calling APIs differing.
It is most likely due to the use of Dask Bags, which are slower than the other Dask APIs.
As with previous experiments, the overhead for Spark is less and scales better, although compensated by a longer total I/O time.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_histogram_worker.jpg}
	\caption{Histogram: varying nodes -- 5000 files, 5 iterations, \SI{1}{\second} delay}
	\label{fig:histogram_worker}
\end{figure}

Figure \ref{fig:histogram_block} shows the execution time for the \textit{Histogram} when varying the number of files.
The results are similar to varying the number of nodes with Spark having faster compute time for \textit{Calculate\_histogram}.
This results in a makespan difference of up to 3\%.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_histogram_block.jpg}
	\caption{Histogram: varying \# of files -- 4 nodes, 5 iterations, \SI{1}{\second} delay}
	\label{fig:histogram_block}
\end{figure}

@@

For brevity we omitted the figure when varying the number of iterations as the results are similar to varying the number of workers.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_kmeans_worker.jpg}
	\caption{Kmeans: varying nodes -- 5000 files, 3 iterations, \SI{1}{\second} delay}
	\label{fig:kmeans_worker}
\end{figure}

@@

\section{Discussion}
\subsection{I/O bottleneck}
We used Lustre as shared file system as it is a \HL{de facto} solution in HPC systems.
The performance of our Lustre installation degrades when a large number of concurrent tasks attempt to perform I/O to the shared file system, as shown in Figures \ref{fig:increment_worker} and \ref{fig:histogram_worker}.
In a previous study~\cite{8943502}, we observed nearly no improvement when increasing the number of nodes to process an application while using an NFS.
\HL{In this paper, the performance scale when increasing the number of nodes, although not linear due to an I/O bottleneck.
	The system was designed such that the filesystem has similar speed to the network.
	Upgrading to more performant networking and storage compononents would reduce the bottleneck seen when increasing the number of workers.

	As in our previous study~{\cite{8943502}}, we observe that the I/O bottleneck has a lower impact on more compute-intensive applications, such as Kmeans and BIDS App Example.
	However, it significantly affects data-intensive applications, such as Increment and Histogram.
}

@@

\subsection{Tradeoff between overhead and data transfer}
Overall, except for Kmeans, the overhead for Dask is larger than for Spark.
However, the total execution time for both engines is comparable.
This almost exact compensation between the engine overhead and the data transfer time (see for example Figure \ref{fig:increment_worker}) is not surprising.
When there is high contention on the bandwidth of the file system, and in the absence of trashing, desynchronizing the file transfers generally does not reduce the total data transfer time.
To take an extreme example, the makespan of $n$ concurrent file transfer of size $F$ from a file system with a bandwidth $B$ would be $nF/B$,
as the $n$ transfer would share the bandwidth $B$, which is equivalent to sequentially transfer the $n$ files.
i.e., when the bandwidth of the file system is saturated, engine overhead does not increase the makespan as long as it desynchronizes the data transfers.
In a previous study~\cite{8943502}, we also observed this tradeoff between overhead and data transfer when the file system was saturated.
It could be interesting to explore strategies that take advantage of this tradeoff to reduce the application makespan.

@@

When reducing the number of processes from 64 to 32 for Spark (effectively doubling the memory of each process), the total execution time became similar between Spark and Dask.
\HL{That is,} with only 32 processes and more memory per worker, Spark no longer recomputed tasks for the Kmeans application.
We suppose that given more memory per thread/process, Dask and Spark would then perform similarly.
\HL{For our experiments, we allocated all available memory on the compute nodes.
Therefore, it was not possible to allocate more resources to test this hypothesis.}

@@

>>>>

We configured Lustre to have 1 MGS and 4 OSS.
Each OSS has 11 OSTs (disks) of \SI{8.8}{\tebi\byte}.
The OSTs are configured to have a maximum of \SI{1000}{\mebi\byte} of dirty data written into the client page cache.
Both the MDT and OSTs are configured to a maximum of 64 concurrent RPCs in flight.
A timeout of \SI{100}{\second} is set for Lustre, and OSTs checksum are disabled.
Lustre \textit{2.14.0-1.el8} is installed on both the Lustre clients and servers.
\HL{We used the default inode configuration for Lustre.}

@@

\subsubsection{Increment}
We adapted the increment application used in~\cite{8752675}.
This synthetic application reads blocks of the BigBrain from Lustre and
simulates computation by sleeping for a specified period (Fig.~\ref{fig:graph-increment}). To simulate
intermediate results, we repeat \HL{a {\SI{1}{\second}} sleep delay during each incrementation}.
We prevent data caching of the blocks by incrementing their voxels
value by one after each sleep operation. Finally, we write the resulting
NIfTI image back to Lustre. This application allows us to study the engines
when their inputs are processed independently. The map-only pattern
mimics the processing of multiple independent subjects in parallel.
% \begin{figure}[!hb]
% 	\centering
% 	\includegraphics[height=0.5\columnwidth,
% 	angle=0]{figures/increment.png}
% 	\caption{Task graph for Incrementation with 3 iterations and 3 BigBrain blocks.}
% 	\label{fig:graph-increment}
% \end{figure}

@@

\HL{Due to resources limitation on our infrastructure, the Kmeans experiments
	are performed with only a sample of half the BigBrain dataset ({\SI{303}{\gibi\byte}}).}

@@

\begin{table*}[t]
	\renewcommand{\arraystretch}{1.5}
	\caption{Parameters for the experiments}\label{table:parameters}
	\centering
	\begin{tabular}{|l|c|c|c|c|c|}
		\hline           & Increment & Histogram & Kmeans & BIDS App Example \\\hline
		Dataset size [\SI{}{\gibi\byte}] &\multicolumn{2}{c|}{606} & 303 & \multicolumn{1}{c|}{26.67} \\\hline
		\# of Nodes & \multicolumn{4}{c|}{2, 4, 8} \\\hline
		Number of file & \multicolumn{2}{c|}{1000, 2500, 5000} & 5000 & \multicolumn{1}{c|}{3491} \\\hline
		\# of Iterations & 1, 5, 25  & -         & 1, 3   & -                \\\hline
	\end{tabular}
\end{table*}

@@

From Figure \ref{fig:increment_worker}, we also observe that Dask's overhead is higher than Spark's.
While the overhead of both engines increases with the number of nodes, Dask's overhead is significantly worse when scaling.
\HL{Although Dask has a higher overhead, the total time for both engines is overall significantly similar.}
This is due to a tradeoff between disk bandwidth and compute/overhead time.
On the one hand, multiple threads access Lustre concurrently, thus sharing the disk bandwidth resulting in longer I/O time.
On the other hand, when most threads are either idle or performing computations, it reduces the contention on the disk bandwidth resulting in shorter I/O time.
\HL{The only significant difference in total time is when two nodes are used.
	This results in an approximately 3\% total runtime difference.}
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_increment_worker.jpg}
	\caption{Increment: varying nodes -- 5000 files, 5 iterations}
	\label{fig:increment_worker}
\end{figure}

@@

Figure \ref{fig:increment_itr} also shows that the the total I/O time decreases as total computation time increases.
This is explained by an I/O bottleneck on the Lustre file system.
When threads are busy with computations, it reduces the contention on Lustre, thus improving I/O speed for individual accesses.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_increment_itr.jpg}
	\caption{Increment: varying iterations -- 4 nodes, 5000 files}
	\label{fig:increment_itr}
\end{figure}

@@

From Figure \ref{fig:increment_block} we also observe that Dask has significantly higher overhead than Spark.
However, the total execution time of the application is \HL{significantly equivalent} due to the tradeoff between I/O and compute/overhead time.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_increment_block.jpg}
	\caption{Increment: varying \# of files -- 4 nodes, 5 iterations}
	\label{fig:increment_block}
\end{figure}

\subsection{Histogram}
Figure \ref{fig:histogram_worker} depicts the execution time for the \textit{Histogram} application when varying the number of nodes.
Overall, \HL{the execution time for Spark is significantly lower than for Dask by up to 5\%}, with the difference coming from a shorter execution time for \textit{Calculate\_histogram}.
This is unexpected as the code base for both engines is the same, with only the calling APIs differing.
It is most likely due to the use of Dask Bags, which are slower than the other Dask APIs.
As with previous experiments, the overhead for Spark is less and scales better, although compensated by a longer total I/O time.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_histogram_worker.jpg}
	\caption{Histogram: varying nodes -- 5000 files}
	\label{fig:histogram_worker}
\end{figure}

Figure \ref{fig:histogram_block} shows the execution time for the \textit{Histogram} when varying the number of files.
The results are similar to varying the number of nodes with Spark having faster compute time for \textit{Calculate\_histogram}.
This results in a \HL{significant} makespan difference of up to 3\%.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_histogram_block.jpg}
	\caption{Histogram: varying \# of files -- 4 nodes}
	\label{fig:histogram_block}
\end{figure}

@@

\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_kmeans_worker.jpg}
	\caption{Kmeans: varying nodes -- 5000 files, 3 iterations}
	\label{fig:kmeans_worker}
\end{figure}

\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=\columnwidth]{figures/dask_kmeans_gantt.png}
	\caption{Kmeans: Dask with 8 nodes, 5000 files, and 3 iterations}
	\label{fig:dask_kmeans_gantt}
\end{figure}

\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=\columnwidth]{figures/spark_kmeans_gantt.png}
	\caption{Kmeans: Spark with 8 nodes, 5000 files, and 3 iterations}
	\label{fig:spark_kmeans_gantt}
\end{figure}

For brevity we omitted the figure\HL{s} when varying the number of iterations as the results are similar to varying the number of workers.

@@

\section{Discussion}
\subsection{I/O bottleneck}
\label{subsec:io-bottleneck}
We used Lustre as shared file system as it is a \HL{de facto} solution in HPC systems.
The performance of our Lustre installation degrades when a large number of concurrent tasks attempt to perform I/O to the shared file system, as shown in Figures \ref{fig:increment_worker} and \ref{fig:histogram_worker}.
In a previous study~\cite{8943502}, we observed nearly no improvement when increasing the number of nodes to process an application while using an NFS.
\HL{In this paper, the performance scale when increasing the number of nodes, although not linear due to an I/O bottleneck.
	The system was designed such that the filesystem has similar speed to the network.
Upgrading to more performant networking and storage compononents would reduce the bottleneck seen when increasing the number of workers.}

\HL{
	As in our previous study~{\cite{8943502}}, we observe that the I/O bottleneck has a lower impact on more compute-intensive applications, such as Kmeans and BIDS App Example.
	However, it significantly affects data-intensive applications, such as Increment and Histogram.
}

@@

\subsection{Tradeoff between overhead and data transfer}
\label{subsec:overhead-tradeoff}
Overall, except for Kmeans, the overhead for Dask is larger than for Spark.
However, the total execution time for both engines is comparable.
This almost exact compensation between the engine overhead and the data transfer time (see for example Figure \ref{fig:increment_worker}) is not surprising.
When there is high contention on the bandwidth of the file system, and in the absence of trashing, desynchronizing the file transfers generally does not reduce the total data transfer time.
To take an extreme example, the makespan of $n$ concurrent file transfer of size $F$ from a file system with a bandwidth $B$ would be $nF/B$,
as the $n$ transfer would share the bandwidth $B$, which is equivalent to sequentially transfer the $n$ files.
i.e., when the bandwidth of the file system is saturated, engine overhead does not increase the makespan as long as it desynchronizes the data transfers.
In a previous study~\cite{8943502}, we also observed this tradeoff between overhead and data transfer when the file system was saturated.
It could be interesting to explore strategies that take advantage of this tradeoff to reduce the application makespan.

@@

When reducing the number of processes from 64 to 32 for Spark (effectively doubling the memory of each process), \HL{the difference in execution time between Spark and Dask reduced significantly}.
\HL{That is,} with only 32 processes and more memory per worker, Spark no longer recomputed tasks for the Kmeans application.
We suppose that given more memory per thread/process, Dask and Spark would then perform similarly.
\HL{For our experiments, we allocated all available memory on the compute nodes.
Therefore, it was not possible to allocate more resources to test this hypothesis.}

@@

\subsection{Limitations}
\HL{
	To make a fair comparison between Spark and Dask, we set the configurations of both
	engine to be equivalent to the best of our capabilities, which comes with some limitations.
	Firstly, both engines use a common implementation for the algorithms.
	This is done to keep the theoretical runtime equivalent, however, in practice the
	chosen implementation might not be optimal for either of the engines.
	Secondly, we decided to use Lustre as a file system for both Spark and Dask.
	Spark could benefit from using HDFS as a file system, however, HDFS is uncommon in HPC.
	Finally, our infrastructure has a bottleneck from the network and Lustre.
	As explained in section~{\ref{subsec:overhead-tradeoff}}, the data transfer time is tighly corrolated to the overhead of the application.
	Therefore, while reducing this bottleneck should improve the performance of both engine similarly
	it can be hard to predict the actual behavior.
}
